Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-04 paper.qmd
Tried to find another good layout for PBMC3k

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_12_min_dist_0.99.rds")
nldr2 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("b")

nldr2c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("b")

@@

nldr3c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour='#f781bf') +
  interior_annotation("c")

@@

nldr4c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("d")

@@

nldr5c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("e")

@@

nldr6c <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("f")

@@

nldr7c <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("g")

@@

nldr8c <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("h")
```

@@

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_15_min_dist_0.99.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_12_min_dist_0.99.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds")
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds")
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

@@

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()
```

@@

error_plot_pbmc <- plot_mse(error_pbmc) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf'))

@@

```{r}
#| label: umap-pbmc-read-best
umap_pbmc2 <- read_rds("data/pbmc3k/pbmc_umap_12_min_dist_0.99.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = umap_pbmc2)
umap_pbmc_scaled_best <- pbmc_scaled_obj$scaled_nldr

@@

## Compute hexbin parameters
num_bins_x_pbmc <- 30

@@

trimesh_removed_pbmc_best <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled_best,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.3,
    color = "#999999"
  )  +
  interior_annotation("b1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

>>>>

# umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_12_min_dist_0.99.rds")
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_3_min_dist_0.6.rds")
nldr2 <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("b")

nldr2c <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("b")

@@

nldr3c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("c")

@@

nldr4c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("d")

@@

nldr5c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("e")

@@

nldr6c <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("f")

@@

nldr7c <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("g")

@@

nldr8c <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#f781bf') +
  interior_annotation("h")
```

@@

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_12_min_dist_0.99.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_3_min_dist_0.6.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds")
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds")
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

@@

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()

error_pbmc <- error_pbmc |>
  mutate(method = factor(method,
                         levels = c("UMAP_3_min_dist_0.6", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))
```

@@

error_plot_pbmc <- plot_mse(error_pbmc) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#377eb8','#4daf4a','#ff7f00','#984ea3','#f781bf'))

@@

```{r}
#| label: umap-pbmc-read-best
umap_pbmc2 <- read_rds("data/pbmc3k/pbmc_umap_3_min_dist_0.6.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = umap_pbmc2)
umap_pbmc_scaled_best <- pbmc_scaled_obj$scaled_nldr

@@

## Compute hexbin parameters
num_bins_x_pbmc <- 29

@@

trimesh_removed_pbmc_best <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled_best,
    aes(
      x = emb1,
      y = emb2
    ),
    alpha = 0.3,
    color = "#999999"
  )  +
  interior_annotation("b1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```
